---
title: "Session 5: Plotting data II"
author: "Jae-Young Son"
output:
  html_document:
    code_folding: show
    toc: TRUE
    toc_float: TRUE
---

This document was most recently knit on `r Sys.Date()`.

# Introduction

In the last tutorial, we covered many of the fundamental ideas underlying the **grammar of graphics**: 1) aesthetics and how to map your data to aesthetics, 2) geometries and how to layer them, and 3) how geometries inherit aesthetics from the initial ggplot call. When you develop a good grasp on these fundamental ideas, you have enough knowledge to figure out how to make nearly any visualization you want (especially now that you know what keywords to search for). In today's tutorial, we'll learn how to give your plots a little more polish. The goal is to make publication-ready plots, which (in theory) won't require further editing in external software like Illustrator or Inkscape.

As you work through this tutorial's material, you may find yourself having to review those fundamental ideas from last time. That's normal, and perfectly fine. You shouldn't expect to become an expert after working through a single tutorial. I'll try to provide some hints and reminders in today's tutorial, which will hopefully let you review and scaffold from your existing knowledge.

Let's go ahead and load some of the same data we were working with last time.

```{r message=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(here)

covid <- here("Data", "time_series_covid19_confirmed_US.csv") %>%
  read_csv() %>%
  clean_names() %>%
  select(-c(uid:fips, country_region:combined_key)) %>%
  rename(county = admin2, state = province_state) %>%
  pivot_longer(cols = -c(county, state), names_to = "date", values_to = "cases") %>%
  mutate(date = str_remove(date, "x"),
         date = mdy(date)) %>%
  group_by(state, date) %>%
  summarise(cases = sum(cases)) %>%
  ungroup()

elections <- here("Data", "countypres_2000-2016.csv") %>%
  read_csv() %>%
  filter(year == 2016) %>%
  filter(party %in% c("democrat", "republican")) %>%
  group_by(state, candidate) %>%
  summarise(candidatevotes = sum(candidatevotes, na.rm=T)) %>%
  group_by(state) %>%
  mutate(lean_democrat = candidatevotes / first(candidatevotes)) %>%
  filter(candidate == "Hillary Clinton") %>%
  ungroup() %>%
  select(state, lean_democrat)

regions <- here("Data", "state_region_division.csv") %>%
  read_csv()
```

# Axis labels

You may have noticed that when you don't supply ggplot with axis labels, it defaults to using your data's variable names:

```{r}
covid %>%
  ggplot(mapping = aes(x=date, y=cases))
```

In this case, it doesn't look terrible, though we probably would prefer the axis labels to be capitalized. However, we'll often be working with variable names that look like `count_normalized_by_pop`. That's considerably less pretty, and I doubt you'd want to publish a figure with that exact axis label. Here's the simplest way to change the axis labels. As a bonus, I've also shown how to add a title and subtitle to your plot.

```{r}
covid %>%
  ggplot(mapping = aes(x=date, y=cases)) +
  xlab("Time") +
  ylab("Raw case numbers") +
  ggtitle(label = "covid-19 cases over time",
          subtitle = "(not normalized by state population)")
```

# Facets

Last time, we explored how to color-code our plots by **mapping** our data to the **aesthetics** `color` and `fill`. That worked okay when we were only comparing a few states to each other, but it was near-impossible to tell colors apart when we were plotting all 50 states (plus DC, territories, and protectorates). This is an example of **overplotting**, i.e. packing so much information into a single plot that it becomes hard to read. Which defeats the whole purpose of graphing out data! To illustrate, let's create a visualization that is clearly overplotted. Try picking out Tennessee from Texas...

```{r fig.height=8}
covid %>%
  ggplot(mapping = aes(x=date, y=cases, color=state)) +
  geom_line() +
  theme(legend.position = "bottom")
```

Since states are discrete categories, it could potentially be useful to separate each state's data out into its own panel. That can be accomplished using a **facet**. In this example, we'll use the function `facet_wrap`. You can see that we specify one argument, `~state`.

The squiggly symbol `~` is called a **tilde**, and can be found (on a US English keyboard at least) above the tab key and to the left of the number 1. When we start learning about statistics, we'll see the tilde very frequently. For example, we might be interested in using a penguin's flipper length to predict its body mass. We would write that **formula** as `body_mass ~ flipper_length`. Colloquially, we might say that we're interested in predicting the penguin's body mass ***by*** its flipper length. The same interpretation holds for our use of `facet_wrap`: we are faceting the data ***by*** state.

```{r fig.height=8}
covid %>%
  ggplot(mapping = aes(x=date, y=cases)) +
  geom_line() +
  facet_wrap(~state) +
  theme(legend.position = "bottom")
```

By looking at the documentation (`?facet_wrap`), you could see how to specify the exact number of columns/rows ggplot generates before it *wraps* the facets (hence the name of the function).

What if we have *two* discrete variables we want to facet by? For example, we could be interested in examining covid counts ***by*** geographical region and political ideology (for illustration, we'll convert the continuous variable `lean_democrat` into a discrete variable). Well, we can take that verbal statement and translate it into a formula, and ggplot will happily accept it.

As you're typing out this code, make sure you can answer the following questions: 1) Why have we added the `group` aesthetic in the initial ggplot call, and what happens if we take it out? 2) Recall that the variable `lean_democrat` is a ratio. Why have we defined the new variable `ideology` in this way? 3) In the previous plot, we faceted by state. So why haven't we specified the formula to be `~state + ideology` in this plot? 4) In our call to `facet_wrap`, why have we specified `ncol = 2`? Why not a different number? 5) What would happen if we made the formula `~ideology + region` instead, and might this change motivate us to specify a different `ncol`?

```{r message=FALSE}
covid %>%
  inner_join(elections) %>%
  mutate(ideology = if_else(
    lean_democrat > 1,
    "Democrat-leaning",
    "Republican-leaning")) %>%
  inner_join(regions) %>%
  ggplot(mapping = aes(x=date, y=cases, group=state)) +
  geom_line() +
  facet_wrap(~region + ideology, ncol = 2)
```

It would be totally fair if you looked at this plot and declared it ugly. When you ask `facet_wrap` to give you multiple facets, it stacks the labels on top of each other. This takes up a lot of space, and doesn't necessarily give you the prettiest output. As an alternative, you could use `facet_grid`:

```{r message=FALSE}
covid %>%
  inner_join(elections) %>%
  mutate(ideology = if_else(
    lean_democrat > 1,
    "Democrat-leaning",
    "Republican-leaning")) %>%
  inner_join(regions) %>%
  ggplot(mapping = aes(x=date, y=cases, group=state)) +
  geom_line() +
  facet_grid(rows = vars(region), cols = vars(ideology))
```

In other people's older code, you might see the syntax `row_variable ~ column_variable` instead. This style is being phased out for three reasons. First, it was hard for people to remember whether **LHS** (left-hand side) corresponded to rows or columns. Second, if you were plotting only a single facet, your code would look like chickenscratch: `. ~ column_variable`. Third, when we were supplying a formula to `facet_wrap`, we understood that `~state` implicitly meant `covid counts ~ state` (or in plain English, covid counts ***by*** state). So in the past, `facet_grid` forced us to specify something that looked like a formula, but was totally inconsistent with our understanding of how formulas work.

```{r message=FALSE}
covid %>%
  inner_join(elections) %>%
  mutate(ideology = if_else(
    lean_democrat > 1,
    "Democrat-leaning",
    "Republican-leaning")) %>%
  inner_join(regions) %>%
  ggplot(mapping = aes(x=date, y=cases, group=state)) +
  geom_line() +
  facet_grid(region ~ ideology)
```

# Scales and legends

We haven't really emphasized this point so far, but every aesthetic has a **scale** associated with it. Let's consider the simplest case of the x and y-axis aesthetics. We're working with a continuous x-axis (time) and a continuous y-axis (case counts). There are two key assumptions we've made when plotting these: 1) that we *do* actually want those scales to be continuous, and 2) that we want the axes to be plotted on a *linear* scale. In the code below, we make this unstated assumption explicit by asking ggplot to use a continuous y-axis scale. While we're at it, we'll also modify the axis title. Recall that we could also do this using `ylab`.

```{r}
covid %>%
  ggplot(aes(x=date, y=cases, group=state)) +
  geom_line() +
  scale_y_continuous(name = "Raw covid-19 case counts")
```

But if we wanted to, we could specify that an aesthetic be represented by a completely different kind of scale. For example, when working with continuous data that has a very large range, it can make sense to plot a **logarithmic** scale instead of a **linear** one.

```{r}
covid %>%
  ggplot(aes(x=date, y=cases, group=state)) +
  geom_line() +
  scale_y_log10(name = "log-transformed covid-19 case counts")
```

Depending on our visualization needs, we could use a variety of scales. Another common scale transformation is to a square-root scale. If (for some reason) we wanted to sort continuous values into bins, we could do that too. Though there's virtually no reason why we would want to visualize these particular data in this way, we could even flip the y-axis scale so that the data are plotted upside down.

```{r}

covid %>%
  ggplot(aes(x=date, y=cases, group=state)) +
  geom_line() +
  scale_y_sqrt(name = "Square-root-transformed covid-19 case counts")

covid %>%
  ggplot(aes(x=date, y=cases, group=state)) +
  geom_line() +
  scale_y_binned(name = "Binned covid-19 case counts")

covid %>%
  ggplot(aes(x=date, y=cases, group=state)) +
  geom_line() +
  scale_y_reverse(name = "Inexplicably-plotted covid-19 case counts")
```

## Discrete color scales

This basic logic extends very naturally to other aesthetics. For example, let's examine color-coding. We could color-code the lines according to geographic region. By default, ggplot knows that region is a **discrete** (i.e., categorical) variable, so it uses a discrete color scale. We can make this assumption explicit by writing the following. Note again that we can change the name of the legend from the variable name `region` by specifying a name of our choosing.

```{r message=FALSE}
covid %>%
  inner_join(regions) %>%
  ggplot(mapping = aes(x=date, y=cases, color=region, group=state)) +
  geom_line() +
  scale_color_discrete(name = "Census-designated region:")
```

If you look at this plot and find it ugly, you are in good company. I've convinced that ggplot's default color palette is deliberately designed to revolt people into specifying their own (better) color scales. More importantly, we should also note that the default palette is not a color palette that is accessible to people with certain forms of color-blindness. Less importantly, if we printed off the plot on a black-and-white printer, we wouldn't be able to tell the colors apart anymore.

So perhaps we'd like to use an alternative color scale that's prettier and more accessible. One built-in option is `viridis`, which is engineered to be perceptually uniform, accessible to viewers with the most common form of color blindness, and convertible to black-and-white. The `d` at the end of `scale_color_viridis_d` refers to the fact that our color aesthetic is being mapped to **discrete** data. There are a couple of viridis palettes: viridis, magma, inferno, plasma, and cividis. You can see that I prefer plasma, and that I even specify a custom subset of the full scale to use (to see why, try removing those arguments). You will develop your own preferences with experience.

```{r message=FALSE}
covid %>%
  inner_join(regions) %>%
  ggplot(mapping = aes(x=date, y=cases, color=region, group=state)) +
  geom_line() +
  scale_color_viridis_d(name = "Census-designated region:",
                        begin = 0.1, end = 0.8, option = "plasma")
```

Another built-in option is `color_brewer`. The best way to understand the brewer is to [play with it online](https://colorbrewer2.org/). Once you've played around with it, the arguments are pretty straightforward to understand:

```{r message=FALSE}
covid %>%
  inner_join(regions) %>%
  ggplot(mapping = aes(x=date, y=cases, color=region, group=state)) +
  geom_line() +
  scale_color_brewer(name = "Census-designated region:",
                     palette = "GnBu")
```

Even though the Brewer is convenient, I don't find myself using it very much. In the plot above, you can hardly see the lines for the Midwest. In practice, I go to the Color Brewer website, pick out colors that I actually like, then feed those **values** to `scale_color_manual`. You might say that the end result below looks bad. Truth be told, I'm actually not the biggest fan of the colors I picked out below. But at the least, I know it's bad because I have poor artistic taste, not because the function automatically picked out "bad" colors on my behalf.

```{r message=FALSE}
covid %>%
  inner_join(regions) %>%
  ggplot(mapping = aes(x=date, y=cases, color=region, group=state)) +
  geom_line() +
  scale_color_manual(name = "Census-designated region:",
                     values = c("#a8ddb5", "#4eb3d3", "#0868ac", "#084081"))
```

## Continuous color scales

So far, we've only worked with discrete color scales. But, the exact same kind of logic works with continuous color scales. For example, let's color-code the lines according to how strongly a state leaned Democrat in the 2016 Presidential election (DC is removed because it overwhelmingly preferred Clinton; including it makes the rest of the plot unreadable):

```{r message=FALSE}
covid %>%
  inner_join(elections) %>%
  filter(state != "District of Columbia") %>%
  ggplot(mapping = aes(x=date, y=cases, color=lean_democrat, group=state)) +
  geom_line() +
  scale_color_continuous(name = "Democrat-voting ratio:")
```

It's often very hard to read continuous color scales. Remember how we'd previously used a y-axis that used a binned scale? That might not be very useful for case counts, but it is often useful for making our color scale more readable.

```{r message=FALSE}
covid %>%
  inner_join(elections) %>%
  filter(state != "District of Columbia") %>%
  ggplot(mapping = aes(x=date, y=cases, color=lean_democrat, group=state)) +
  geom_line() +
  scale_color_binned(name = "Democrat-voting ratio:")
```

You can modify the palette of continuous color scales in (more-or-less) exactly the same way we modified discrete color scales. On your own, try using viridis, brewer, and manual color scales.

## Other aesthetics

I won't spend much time on this section. I only want to note that there are similar scales for other aesthetics, and reiterate that every aesthetic is represented by a scale even when it's not explicitly specified. ggplot is usually pretty good about picking default scales, but just remember that you can take complete control at any time, for any aesthetic.

# Coordinates

In your history or math classes, you might remember having learned about the ancient Greek mathematician Euclid. His major accomplishment was deriving a system for describing geometric objects in terms of a small number of fundamental assumptions (known as axioms), which could be logically combined into theorems (statements that are true, as long as the axioms are true and have been combined without error). You may have heard the term "Euclidean geometry" used to describe this system of thinking.

When we plot data, we are interested in mapping those data to geometries, and we have already spent considerable time considering this idea. But so far, we've somewhat taken for granted that these geometries also exist within a **coordinate system**. Let's imagine that we draw a square on a balloon that hasn't been inflated yet. We can describe the properties of that square quite well using Euclidean geometry. Then, we blow air into the balloon until it's fully inflated. When we look at the drawing we've made, will it still look like a square? Not really, because we've warped the underlying space. And if we tried to use Euclidean geometry to analyze that drawing, we wouldn't get very far.

For a moment, let's return to the x/y-axis scales. These are special because they modify the **coordinate system** of the entire plot. For example, if we use `scale_y_sqrt` to plot the data on a square-root-transformed y-axis, we're no longer operating in a linear coordinate space.

Most of the time, when we need to modify the x/y-axis scales, it's easiest to do it through `scale_x_whatever` or `scale_y_whatever`. However, there are going to be a few times when you'll want to modify the coordinate system

# Themes

# Exercises
